---
author: Daniel Wysocki
title: Markov Processes
date: March 12, 2015
output:
  beamer_presentation:
    toc: true
    theme: Madrid
    colortheme: dolphin
    fonttheme: serif
    includes:
      in_header: [markov-processes-header.tex]
---

```{r echo=FALSE, message=FALSE}
require(knitcitations)
cleanbib()
bib <- read.bibtex("markov-processes.bib")
```

# Stochastic Processes

## Definition

A stochastic process is a collection of random variables $\{ X(t), t \in T \}$
defined on a common probability space indexed by the index set $T$ which
describes the evolution of some system.

`r citep(bib[["resnick1992adventures"]])`


## Properties

- the evolution of the system is non-deterministic, even if all initial
  variables are known

- the system may evolve in many (possibly infinite) ways

- any given sequence of events may be assigned a probability

- the probability of all possible sequences must sum to 1


## History

- early motivation for the theory of stochastic processes lies in Brownian
  motion

- Brownian motion was a phenomenon observed by Robert Brown in 1827

    - he observed under a microscope the motions of pollen in a fluid

    - the movements seemed unpredictable

- Brownian motion is the outcome of many unpredictable or unobservable events,
  each imparting a negligible influence on an observed phenomenon, but
  collectively having a significant influence

`r citep(bib[["paul2013stochastic"]])`


## History (cont.)

- Louis Bachelier made his PhD thesis in 1900, applying stochastic theory to
  the price of financial assets

    - the significance of his work was not realized until much later

    - it contained many of the results of stochastic theory used today

    - his advisor, Henri Poincaré, when trying to solve the Brownian motion
      problem years later, did not realize Bachelier had already solved it

`r citep(bib[["paul2013stochastic"]])`


## History (cont.)

- in his 1906 article in *Annals of Physics*, Albert Einstein described the
  mechanisms which drive Brownian molecular motion

- this helped strengthen the evidence for atoms and molecules

- presented a method for measuring the size of the atom

`r citep(bib[["einstein1956investigations"]])`


## Markov Processes

- a special case of stochastic processes

- satisfy the property that the future is dependent only on the present,
  not the past

- may not be a perfect model of the underlying stochastic process, but often
  serves as a good approximation, and reduces the complexity of the model

`r citep(bib[["resnick1992adventures"]])`


## History

- Andrey Markov, a Russian mathematician, developed a technique, now known as
  the Markov chain

- he used it to discover patterns in the vowels and consonants of Alexander
  Pushkin's novel, *Eugene Onegin*

    - published a summary of his findings in an address to the Imperial
      Academy of Sciences in St. Petersberg in 1913

- though his work did little to understand the works of Pushkin, it started a
  new branch of probability theory, in wide use today

`r citep(bib[["hayes2013first"]])`


# Markov Chains

## Definition

- Markov chains are Markov processes with a discrete index set $T$, and a
  countable or finite state space

- the Markov chain can be described by only two features: a 1-dimensional
  initial probability distribution, $a(k)$, and a 2-dimensional transition
  probability distribution, $p(i, j)$

    - $a(k)$ is the probability that the first element of the chain is in state
      $k$

    - $p(i, j)$ is the probability that a system in state $i$ will transition
      into $j$

$$\sum_{k=1}^N a(k) = \sum_{j=1}^N p(i, j) = 1,$$
where $N$ is the number of possible states

`r citep(bib[["resnick1992adventures"]])`


## Graph Representation

A simple Markov process, with two states $A$ and $B$, can be represented by
this graph

\begin{figure}[H]
\centering
\markovchain
\end{figure}


## Training a Model

- a Markov chain may be constructed by observing a sequence of events, and
  constructing a transition frequency matrix and an initial frequency vector

- then we normalize the two, such that the rows sum to 1

    - row stochastic

- consider the observed sequences $(A, A, B, A, B)$ and $(B, A, B, B, A)$

$$
\tilde{a} = \begin{bmatrix} 1 & 1 \end{bmatrix},
\quad
\tilde{p} = \begin{bmatrix} 1 & 3 \\ 1 & 3 \end{bmatrix}
\implies
a = \begin{bmatrix} 1/2 & 1/2 \end{bmatrix},
\quad
p = \begin{bmatrix} 1/4 & 3/4 \\ 1/4 & 3/4 \end{bmatrix}
$$


## Generating Sequences

- once we have trained a model, we may use it to generate random sequences,
  which we shall denote as $X$

- we utilize a random number generator, and with probability $a(k)$ we choose
  initial state $k$, so $X_0 = k$

- subsequent states can be found recursively, such that we choose state
  $X_n$ with probability $p(X_{n-1}, X_n)$


## Graphical Sequence Generation

We can envision the process of generating a sequence as traversing this graph.
At each node, we use the associated transition probabilities to decide where
to go next

\begin{figure}[H]
\centering
\markovchain
\end{figure}


# Hidden Markov Models

## Definition

In a hidden Markov model, the states in the Markov chain are not directly
observable. However, there is a set of observables variables, which are
somehow correlated with the underlying (or hidden) Markov chain.

`r citep(bib[["alpaydin2010introduction"]])`


## History

- early theory of HMMs developed by Leonard E. Baum and colleagues at the
  Institute for Defense Analyses (IDA) in the 1960's

- Jim Baker applied HMMs to speech recognition in the 1970's, but was not
  100% successful

- Jack Ferguson and colleagues at IDA gave a classical series of lectures in
  1980, which boosted the popularity of HMMs

- continued usage in speech recognition

- now widely used in biological sequencing

`r citep(list(bib[["rabinerIEEE"]], bib[["yoon2009hidden"]]))`


## Graph Representation

\begin{figure}[H]
\centering
\hiddenmarkovmodel
\end{figure}


## Parameters

- a hidden Markov process can be parameterized by the following variables

    - the states $S = \{ S_1, S_2, \ldots, S_N \}$

    - the observation symbols $V = \{ v_1, v_2, \ldots, v_M \}$

    - the state transition probabilities
      $\mathbf{A} = [a_{ij}]
       \text{, where }
       a_{ij} = P(q_{t+1} = S_j | q_t = S_i)$

    - the observation probabilities
      $\mathbf{B} = [b_j(m)]
       \text{, where }
       b_j(m) = P(O_t = v_m | q_t = S_j)$

    - the initial state probabilities
      $\boldsymbol\Pi = [\pi_i]$ where $\pi_i = P(q_1 = S_i)$

- the model itself is denoted
  $\lambda = (\mathbf{A}, \mathbf{B}, \boldsymbol\Pi)$

- the observation sequence is denoted $O = ( O_1, O_2, \ldots, O_T )$

- the hidden state sequence is denoted $Q = ( q_1, q_2, \ldots, q_T )$


## Traffic Light Markov Chain

Suppose we want to model a traffic light as a Markov process. This can be done
using a very simple Markov chain.

There are 3 states,
\textcolor{red}{red}, \textcolor{yellow}{yellow}, and \textcolor{green}{green}.
Each state has 2 non-zero transition probabilities, either it remains the same,
or changes in the progression
\textcolor{green}{green} $\to$
\textcolor{yellow}{yellow} $\to$
\textcolor{red}{red} $\to$
\textcolor{green}{green} ...

\begin{figure}[H]
\centering
\begin{tikzpicture}[-latex, auto, node distance = 3cm and 3cm, on grid,
                    semithick,
                    state/.style = {circle,
                                    top color = white,
                                    bottom color = black!80,
                                    draw,
                                    black,
                                    text = black,
                                    minimum width = 1.5 cm}]

\node[state] (R) [bottom color = red] {Red};
\node[state] (Y) [bottom color = yellow, right = of R] {Yellow};
\node[state] (G) [bottom color = green, right = of Y] {Green};

\path (G) edge [loop right] (G);
\path (Y) edge [loop above] (Y);
\path (R) edge [loop left]  (R);

\path (G) edge [bend right = 15] (Y);
\path (Y) edge [bend right = 15] (R);
\path (R) edge [bend right = 25] (G);

\end{tikzpicture}
\end{figure}


## Traffic Light HMM

Suppose we want to model this same traffic light, but we only have a video
stream from a satellite. The satellite cannot see the light itself, but can
resolve the velocity of the moving cars, both in the same direction as the
light, and the intersecting direction.

The hidden states are
\textcolor{red}{red}, \textcolor{yellow}{yellow}, and \textcolor{green}{green}.
The observables are
fast same direction ($\Updownarrow$),
slow same direction ($\updownarrow$),
fast intersecting direction ($\Leftrightarrow$),
slow intersecting direction ($\leftrightarrow$),
all stopped ($\otimes$).

Assuming cars don't run red lights, the HMM might look something like.

\begin{figure}[H]
\centering
\begin{tikzpicture}[-latex, auto, node distance = 1.5cm and 1.5cm, on grid,
                    semithick,
                    state/.style = {circle,
                                    top color = white,
                                    bottom color = black!80,
                                    draw,
                                    black,
                                    text = black,
                                    minimum width = 0.5 cm},
                    obs/.style   = {rectangle,
                                    top color = white,
                                    bottom color = blue!40,
                                    draw,
                                    blue,
                                    text = black,
                                    minimum width = 0.5 cm}]

\node[state] (R) [bottom color = red] {R};
\node[state] (Y) [bottom color = yellow, right = of R] {Y};
\node[state] (G) [bottom color = green, right = of Y] {G};

\node[obs] (X) [below = of Y] {$\otimes$};
\node[obs] (s) [right = of X] {$\updownarrow$};
\node[obs] (S) [right = of s] {$\Updownarrow$};
\node[obs] (i) [left = of X] {$\leftrightarrow$};
\node[obs] (I) [left = of i] {$\Leftrightarrow$};

\path (G) edge [loop right] (G);
\path (Y) edge [loop above] (Y);
\path (R) edge [loop left]  (R);

\path (G) edge [bend right = 15] (Y);
\path (Y) edge [bend right = 15] (R);
\path (R) edge [bend right = 25] (G);

\path (R) edge [bend right = 15, color = blue] (I);
\path (R) edge [color = blue] (i);
\path (R) edge [bend left = 15, color = blue] (X);

\path (Y) edge [color = blue] (X);
\path (Y) edge [bend right = 15, color = blue] (s);
\path (Y) edge [bend left = 15, color = blue] (S);

\path (G) edge [bend right = 15, color = blue] (X);
\path (G) edge [color = blue] (s);
\path (G) edge [bend left = 15, color = blue] (S);

\end{tikzpicture}
\end{figure}


## Common HMM Problems

Some common problems occur in HMMs, which have standard solutions.

1. Given the model, $\lambda$, and a sequence of observations, $O$,
   determine the likelihood of observing that sequence, $P(O | \lambda)$

    - this is solved by the forward algorithm
    - might be used to detect unusual traffic

2. Given the model, $\lambda$, and a sequence of observations, $O$,
   find an optimal state sequence $Q$

    - this is solved by the backward algorithm
    - this will tell us the most likely sequence of traffic light states

3. Given an observation sequence, $O$, and the dimensions of the model,
   $N$ and $M$, find the model $\lambda$ that maximizes the probability of
   $O$

    - this is solved by the Baum--Welch algorithm
    - this is how we initially train the model

`r citep(bib[["stamp2012"]])`


## Forward Algorithm

1. Let $\alpha_0(i) = \pi_i b_i(O_0)$, for $i = 0, 1, \ldots, N-1$

2. For $t = 1, 2, \ldots, T-1$ and $i = 0, 1, \ldots, N-1$, compute
$$\alpha_t(i) = \left[
  \sum_{j=0}^{N-1} \alpha_{t-1}(j) a_{ji}
\right] b_i(O_t)$$

3. Since $\alpha_t(i) = P(O_0, O_1, \ldots, O_t, x_t=q_i | \lambda)$,
   it follows that
$$P(O|\lambda) = \sum_{i=0}^{N-1} \alpha_{T-1}(i)$$

`r citep(bib[["stamp2012"]])`


## Backward Algorithm

1. Let $\beta_{T-1}(i) = 1$, for $i = 0, 1, \ldots, N-1$.

2. For $t = T-2, T-3, \ldots, 0$ and $i = 0, 1, \ldots, N-1$, compute
$$\beta_t(i) = \sum_{j=0}^{N-1} a_{ij} b_j(O_{t+1}) \beta_{t+1}(j)$$

3. For $t = 0, 1, \ldots, T-2$ and $i = 0, 1, \ldots, N-1$, define
   $\gamma_t(i) = P(x_t = q_i | O, \lambda)$,
   which can be written as
   $$\gamma_t(i) = \frac{\alpha_t(i) \beta_t(i)}{P(O | \lambda)}$$

The most likely state at time $t$ is the state $q_i$ with maximum $\gamma_t(i)$.

`r citep(bib[["stamp2012"]])`


## Baum--Welch Algorithm

- define "di-gamma" as
$$
\gamma_t(i, j) =
\frac{\alpha_t(i) a_{ij} b_j(O_{t+1}) \beta_{t+1}(j)}{P(O | \lambda)}
$$
- this is related to gamma by
$$\gamma_t(i) = \sum_{j=0}^{N-1} \gamma_t(i, j)$$
- given $\gamma$ and di-gamma, we can re-estimate a model using the following
  procedure

`r citep(bib[["stamp2012"]])`


## Baum--Welch Algorithm

1. for $i = 0, 1, \ldots, N-1$, let
$$\pi_i = \gamma_0(i)$$
2. for $i = 0, 1, \ldots, N-1$ and $j = 0, 1, \ldots, N-1$, compute
$$
a_{ij} =
\frac{\sum_{t=0}^{T-2} \gamma_t(i, j)}{\sum_{t=0}^{T-2} \gamma_t(i)}
$$
3. for $j = 0, 1, \ldots, N-1$ and $k = O_0, O_1, \ldots, O_{M-1}$, compute
$$
b_j(k) =
\frac{\sum_{\substack{t \in \{0, 1, \ldots, T-2 \} \\ O_t = k}} \gamma_t(j)}
     {\sum_{t=0}^{T-2} \gamma_t(j)}
$$

`r citep(bib[["stamp2012"]])`


## Baum--Welch Algorithm

Now that $\gamma_t(i, j)$ and re-estimation have been defined, we can write
the entire algorithm as

1. initialize $\lambda = (A, B, \pi)$.

2. compute $\alpha_t(i)$, $\beta_t(i)$, $\gamma_t(i, j)$ and $\gamma_t(i)$

3. re-estimate the model according the the method just defined

4. if $P(O | \lambda)$ increases by some threshold, or max iterations reached,
   repeat from step 2

`r citep(bib[["stamp2012"]])`


## Hidden Markov Music

- MIDI events (or some invertible function of them) are the observations

- a musical piece is an observation sequence $O$

- a model, $\lambda$, is trained on some set of musical pieces, $\mathbf{O}$

- the hidden states are completely determined by the observations, and might
  correspond to anything

    - the only human intervention is in choosing the number of states, $N$

- the forward algorithm can be used to find $P(O | \lambda)$ for a given piece,
  which may be interpreted as how well it matches the model (classification)

- new pieces may be generated from the model, by walking through the model,
  making choices based on transition and emission probabilities


---

\small

```{r, echo=FALSE, results='asis'}
bibliography()
```
